---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Housekeeping

Clear Environment and Console

```{r clear_environment, echo=TRUE, results='hide'}
rm(list=ls())
cat('\014')
```

packageR is a function used to install R Packages.<br />
packageR is invoked, and packages installed, only when needed

```{r packageR, echo=TRUE, results='hide'}
packageR <- function(pkg){
     if (!require(pkg,character.only = TRUE))
     {
          install.packages(pkg,dep=TRUE)
          if(!require(pkg,character.only = TRUE)) {
               stop("Package not found")
          } else {
               library(pkg,character.only=TRUE)
          }
     }
}

packageR("datasets")
packageR("ProjectTemplate")
packageR("ggplot2")
packageR("lattice")
packageR("scales")
packageR("rmarkdown")
packageR("knitr")
packageR("data.table")
packageR("dplyr")
packageR("lubridate")
packageR("sqldf")
packageR("Hmisc")
```

pathR is a function to simplify multi-device development.<br />
pathR is used to set 'path_to_data' as appropriate for my machines

```{r pathr, echo=TRUE, results='hide'}
pathR <- function(machine,course){
     machine.path <- "/Users/rickwrites/Dropbox/Data-Science/Coursera/"
     if(machine != "iMac"){
          machine.path <- "/Users/Rick/Dropbox/Data-Science/Coursera/"
     }
     paste(machine.path,course,sep="")
}

getwd()
course <- "DS-05-Reproducible/RepData_PeerAssessment1"
working.directory <- pathR('iMac',course)
setwd(working.directory) ## Make it so, Number One (set the Working Directory to the applicable path for current machine)
getwd()

```

Configure key parameters for knitr

```{r setup_knitr_params, echo=TRUE}
opts_chunk$set(echo = TRUE, cache = TRUE, cache.path = "cache/", fig.path = "figure/")
```

## Loading and preprocessing the data<br />
1. Load the data (<b>Note:</b> Because <b>data.table</b> is being used, fread, not read.csv, is used to populate the dataframe)

```{r load_data, echo=TRUE}
csv.name <- "activity.csv"

if (!file.exists(csv.name)){
     unzip("activity.zip")
}

dt.activity <- fread("activity.csv",na.strings=(c("NA")))
```

2. Process/transform the data (if necessary) into a format suitable for your analysis<br />
2.a Coerce 'date' from Chr into Date data type (interesting finding: POSIX dates don't work)<br />
2.b Coerce 'steps' from Int into Numeric to ensure it works with median function, later

```{r conversions, echo=TRUE}
dt.activity$date <- as.Date(dt.activity$date)
dt.activity$steps <- as.numeric(dt.activity$steps)
```

Verify data.table and conversions

```{r verify.01, echo=TRUE}
print(str(dt.activity))
print(nrow(dt.activity))
head(dt.activity,20)
```

## What is mean total number of steps taken per day?
Note from Instructions: <i>"...ignore the missing values in the dataset"</i><br />
1. Calculate the total number of steps taken per day (this can be accomplished with 'stat_summary', as shown below)<br />
2. Make a histogram of the total number of steps taken each day

```{r histogram_plot_count_steps_per_day_ignore_na, echo=TRUE}
p.1 <- ggplot(data = na.exclude(dt.activity), aes(x = date, y = steps))
p.1 <- p.1 + stat_summary(fun.y = sum, geom = "histogram")
p.1 <- p.1 + ggtitle("Count of Steps by Date (Excluding Missing Values)") + ylab("Count of Steps") + xlab("Date")

print(p.1)
```

3. Calculate and report the mean and median of the total number of steps taken per day<br />
<b>IMPORTANT NOTE:</b> Because '0' steps occur so frequently in the source data--which means the median would be '0'--<br />
This section of the assignment is being conducted with <b>TWO</b> datasets:<br />
One that <b>includes</b> '0' values in mean & median computations, and<br />
A second that <b>excludes</b> '0' values in mean & median computations

First, mean & median with zeroes included

```{r compute_mean_of_total_steps_by_date, echo=TRUE}
total.steps.by.date <- na.exclude(dt.activity[,lapply(.SD,sum,na.rm=TRUE),by=date,.SDcols='steps'])

mean.of.total.steps.by.date <- na.exclude(dt.activity[,lapply(.SD,mean,na.rm=TRUE),by=date,.SDcols='steps'])
setnames(mean.of.total.steps.by.date,c("date","steps"), c("date","mean_of_steps"))
print(mean.of.total.steps.by.date)

median.of.total.steps.by.date <- na.exclude(dt.activity[,lapply(.SD,median,na.rm=TRUE),by=date,.SDcols='steps'])
setnames(median.of.total.steps.by.date,c("date","steps"), c("date","median_of_steps"))
print(median.of.total.steps.by.date)
```

Second, mean & median <b>without</b> zeroes (zeroes excluded)

```{r compute_median_of_total_steps_by_date, echo=TRUE}
activity.zeros.as.na.dt <- dt.activity
activity.zeros.as.na.dt[activity.zeros.as.na.dt$steps == 0] <- NA

mean.of.total.steps.by.date.exclude.zeroes <- na.exclude(activity.zeros.as.na.dt[,lapply(.SD,mean,na.rm=TRUE),by=date,.SDcols='steps'])
setnames(mean.of.total.steps.by.date.exclude.zeroes,c("date","steps"), c("date","mean_of_steps"))
print(mean.of.total.steps.by.date.exclude.zeroes)

median.of.total.steps.by.date.exclude.zeroes <- na.exclude(activity.zeros.as.na.dt[,lapply(.SD,median,na.rm=TRUE),by=date,.SDcols='steps'])
setnames(median.of.total.steps.by.date.exclude.zeroes,c("date","steps"), c("date","median_of_steps"))
print(median.of.total.steps.by.date.exclude.zeroes)
```

Although the assignment does not specify if the mean & median are to be plotted, or not--<br />
Just to be sure: a plot of both the mean & median of total steps by day are included for completness

```{r plot_mean_median_of_total_steps_by_date, echo=TRUE}
p.mean <- ggplot(data = na.exclude(mean.of.total.steps.by.date.exclude.zeroes), aes(x = date, y = mean_of_steps))
p.mean <- p.mean + geom_line()
p.mean <- p.mean + ggtitle("Line Plot of Means of Total Number of Steps by Date - Excluding Zero Values") + ylab("Mean of Daily Steps (Zeroes Excluded") + xlab("Date")
print(p.mean)

p.median <- ggplot(data = na.exclude(median.of.total.steps.by.date.exclude.zeroes), aes(x = date, y = median_of_steps))
p.median <- p.median + geom_line()
p.median <- p.median + ggtitle("Line Plot of Medians of Total Number of Steps by Date - Excluding Zero Values") + ylab("Median of Daily Steps (Zeroes Excluded") + xlab("Date")
print(p.median)
```

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)<br />
<b>IMPORANT NOTE:</b> The 'interval' varable is in the form of [#]hmm; that is: [0]155 is Hour 1, Minute 55; which is followed by [0]200, or Hour 2, Minute 00.,br />
First step is to use the <b>sqldf</b> package to group the active data.table by the interval identifier, then average it

```{r average_activity_pattern, echo=TRUE}
sql.select <- "select interval, avg(steps) as avg_steps from 'dt.activity' group by interval"
dt.avg.activity.pattern <- sqldf(sql.select)
print(head(dt.avg.activity.pattern))
```

Now, prepare and print time-series plot (type="l") using <b>xyplot</b> from lattice package

```{r time_series_plot_avg_activity_pattern, echo=TRUE}
xyplot(avg_steps ~ interval,
       data = dt.avg.activity.pattern,
       type = "l",
       main = "Average Daily Activity Pattern Grouped by Interval Identifier",
       xlab = "Interval Identifer",
       ylab = "Average of Steps Grouped by Interval")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r determine_max_interval, echo=TRUE}
max.steps.interval <- dt.avg.activity.pattern[max(dt.avg.activity.pattern$avg_steps),]
print(max.steps.interval)
```
The 5-minute interval that has the maximum number of steps (averaged across all days in the dataset) is interval id <b>`r max.steps.interval$interval`</b>, which reflected a maximum value of <b>~`r format(max.steps.interval$avg_steps,digits=3)`</b>.

## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA).<br />
The presence of missing days may introduce bias into some calculations or summaries of the data.<br /><br />
1. Calculate and report the total number of missing values in the dataset (i.e.. the total number of rows with NAs).<br />
In the source dataset, there is a total of <b>`r nrow(dt.activity)`</b> rows; of which <b>`r sum(is.na(dt.activity))`</b> have missing values.<br /><br />
2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.<br /><br />
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.<br /><br />
### Approach to Transform Imputed Values
The approach taken to ransform imputed value is:<br />
1. Observations whose original 'steps' value is zero ('0') <b>remain as zero</b>, and<br />
2. <b>'NA'</b> observations are transformed to be the rounded value of the 'avg_steps' (with a minimum value of '1') for the corresponding interval, as computed in a prior section (above)<br />
3. Create a new data.table that reflects the modified imputed values (the <b>new data.table is 'merged_all'</b>, as shown below)

```{r set_na_to_avg_steps, echo=TRUE}
merge_x <- as.data.table(dt.activity)
setkey(merge_x,interval)
merge_y <- as.data.table(dt.avg.activity.pattern)
setkey(merge_y,interval)

merged <- merge(merge_x,merge_y,by.x = interval, by.y = interval)
needsfill <- merged[is.na(merged$steps)]
needsfill$steps <- needsfill$avg_steps
needsfill$steps <- round(needsfill$steps)

merged <- merge(merge_x,merge_y,by.x = interval, by.y = interval)

merged_0 <- merge(merge_x,merge_y,by.x = interval, by.y = interval)

obs_nominal_1 <- merged_0[!is.na(merged_0$steps)]
obs_fillna <- merged_0[is.na(merged_0$steps)]
obs_fillna$steps <- round(obs_fillna$avg_steps)

obs_nominal_2 <- obs_fillna[obs_fillna$steps >= 1]
obs_fillzeroes <- obs_fillna[obs_fillna$steps < 1]
obs_fillzeroes$steps <- 1
merged_fill <- rbind(obs_nominal_2,obs_fillzeroes)
merged_all <- rbind(obs_nominal_1,merged_fill)
```

Now, inspect the results as contained in the 'merged_all' data.table

```{r verify.03, echo=TRUE}
print(head(merged_all))
print(str(merged_all))
```

4.a Make a histogram of the total number of steps taken each day.

```{r histogram_plot_count_steps_per_day_with_imputed_values, echo=TRUE}
p.2 <- ggplot(data = merged_all, aes(x = date, y = steps))
p.2 <- p.2 + stat_summary(fun.y = sum, geom = "histogram")
p.2 <- p.2 + ggtitle("Count of Steps by Date (Missing Values Imputed") + ylab("Count of Steps") + xlab("Date")

print(p.2)
```

4.b  Calculate and report the mean and median total number of steps taken per day [Using the <i>imputed dataset</i>].<br />

```{r compute_mean_of_imputed_total_steps_by_date, echo=TRUE}
total.imputed.steps.by.date <- merged_all[,lapply(.SD,sum,na.rm=TRUE),by=date,.SDcols='steps']

mean.of.imputed.total.steps.by.date <- merged_all[,lapply(.SD,mean,na.rm=TRUE),by=date,.SDcols='steps']
setnames(mean.of.imputed.total.steps.by.date,c("date","steps"), c("date","mean_of_steps_imputed"))
print(mean.of.imputed.total.steps.by.date)

median.of.imputed.total.steps.by.date <- merged_all[,lapply(.SD,median,na.rm=TRUE),by=date,.SDcols='steps']
setnames(median.of.imputed.total.steps.by.date,c("date","steps"), c("date","median_of_steps_imputed"))
print(median.of.imputed.total.steps.by.date)
```

Do these values differ from the estimates from the first part of the assignment?<br />
After the imputation process, there is a significant difference in the resulting datasets.<br />
In the original mean & median analysis (above)&mdash;when "NAs" were excluded&mdash;there were <b>`r nrow(mean.of.total.steps.by.date)`</b> observations in the dataset.<br />
After the data are modified to impute missing values, the resultant mean & median analysis reflects <b>`r nrow(mean.of.imputed.total.steps.by.date)`</b> observations.

```{r combine_source_and_imputed_mean_and_median_datasets, echo=TRUE}
setkey(mean.of.total.steps.by.date, date)
setkey(mean.of.imputed.total.steps.by.date, date)
combined_means <- mean.of.total.steps.by.date[mean.of.imputed.total.steps.by.date]

setkey(median.of.total.steps.by.date, date)
setkey(median.of.imputed.total.steps.by.date, date)
combined_medians <- median.of.total.steps.by.date[median.of.imputed.total.steps.by.date]
```

```{r compute_mean_median_differences, echo=TRUE}
combined_means$meandiff <- (combined_means$mean_of_steps - combined_means$mean_of_steps_imputed)
print(combined_means)

combined_medians$mediandiff <- (combined_medians$median_of_steps - combined_medians$median_of_steps_imputed)
print(combined_medians)
```

<b>ANALYSIS RESULTS:</b><br />
Mean: `r nrow(combined_means[combined_means$meandiff > 0])` days were affected by imputation.<br />
Median: `r nrow(combined_medians[combined_medians$mediandiff > 0])` days were affected by imputation.<br />


What is the impact of imputing missing data on the estimates of the total daily number of steps?


## Are there differences in activity patterns between weekdays and weekends?


